---
title: "eda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)   
library(skimr)        
library(janitor)      
```

## Data Import

First, we import the dataset from a local CSV file. The clean_names() function from janitor ensures that variable names are standardized.

```{r}
energy_raw <- read_csv("owid-energy-data.csv")

#Make column names consistent
energy <- clean_names(energy_raw)

head(energy)
```

## Understanding the Dataset and Variables

Here we use skimr to get a quick summary of all variables.

```{r}
#dim(energy)

#glimpse(energy)

skim(energy)

# This can be used to see some basic summaries for numerical variables 
#summary(select(energy, where(is.numeric)))
```

## Selection

Since we have 129 columns, let's focus on ones potentially relevant to our research questions:

```{r}
eda_df <- energy %>%
  select(
    country, year, iso_code, population, gdp,
    primary_energy_consumption, fossil_fuel_consumption,
    coal_consumption, gas_consumption,
    renewables_electricity, solar_electricity,
    wind_electricity, hydro_electricity
  )
```


## Basic Summaries & Missingness

```{r}
summary(eda_df)

# checking missing values
eda_df %>%
  summarize_all(~ sum(is.na(.))) %>%
  gather(key = "column_name", value = "num_missing") %>%
  arrange(desc(num_missing))
```


## Distribution Plots

We plot the distribution of primary energy consumption.

```{r}
eda_df %>%
  filter(!is.na(primary_energy_consumption)) %>%
  ggplot(aes(x = primary_energy_consumption)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black", alpha = 0.7) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Primary Energy Consumption",
    x = "Primary Energy Consumption (TWh)",
    y = "Frequency"
  ) +
  theme_minimal()

```

However, the scale is not optmial so a better way would be to use log-scale

```{r}
eda_df %>%
  filter(!is.na(primary_energy_consumption) & primary_energy_consumption > 0) %>%
  ggplot(aes(x = log10(primary_energy_consumption))) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Log(Primary Energy Consumption)",
    x = "Log10(Primary Energy Consumption)",
    y = "Frequency"
  ) +
  theme_minimal()
```


## Checking Correlation


We can get an idea about the data if we check the correlation between GDP and Primary Energy Consumption

```{r}
eda_df %>%
  filter(!is.na(gdp), !is.na(primary_energy_consumption), # Filtering out missing values to avoid infinite or meaningless logs
         gdp > 0, primary_energy_consumption > 0) %>%
  ggplot(aes(x = gdp, y = primary_energy_consumption)) +
  geom_point(alpha = 0.3, color = "blue") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  labs(
    title = "GDP vs Primary Energy Consumption",
    x = "GDP",
    y = "Primary Energy Consumption (TWh)"
  )
```

However, log-tranforming can give us a better graph in this case too

```{r}
eda_df %>%
  filter(!is.na(gdp), !is.na(primary_energy_consumption),
         gdp > 0, primary_energy_consumption > 0) %>%
  ggplot(aes(x = log10(gdp), y = log10(primary_energy_consumption))) +
  geom_point(alpha = 0.3, color = "blue") +
  theme_minimal() +
  labs(
    title = "Log-Log Plot: GDP vs. Primary Energy Consumption",
    x = "Log10(GDP)",
    y = "Log10(Primary Energy Consumption)"
  )
```

## Time Trends

We look at how primary energy consumption changes over time for the top gdp countries in the world

```{r}
top_countries <- eda_df %>%
  filter(!is.na(gdp), country != "World") %>%  
  group_by(country) %>%                       
  summarize(total_gdp = sum(gdp, na.rm = TRUE)) %>% 
  arrange(desc(total_gdp)) %>%                
  slice_head(n = 5) %>%                      
  pull(country)                               

eda_df %>%
  filter(country %in% top_countries,
         !is.na(primary_energy_consumption)) %>%
  ggplot(aes(x = year, y = primary_energy_consumption, color = country)) +
  geom_line() +
  labs(
    title = "Primary Energy Consumption Over Time",
    x = "Year",
    y = "Primary Energy Consumption (TWh)"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```
We can also explore Renewables (e.g., wind, solar, hydro) Over Time for each of the selected country

```{r}
eda_df %>%
  filter(country %in% top_countries) %>%
  select(country, year, renewables_electricity, solar_electricity,
         wind_electricity, hydro_electricity) %>%
  pivot_longer(
    cols = c(renewables_electricity, solar_electricity, wind_electricity, hydro_electricity),
    names_to = "renewable_type",
    values_to = "electricity_twh"
  ) %>%
  filter(!is.na(electricity_twh)) %>%
  ggplot(aes(x = year, y = electricity_twh, color = renewable_type)) +
  geom_line() +
  facet_wrap(~ country, scales = "free_y") +
  labs(
    title = "Trends in Different Renewable Electricity Sources",
    x = "Year",
    y = "Electricity (TWh)"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```


## Trends (Part2)

Question: Do countries that exhibit rapid population growth also show significant increases in overall energy demand?

Filter the data for the years and for non-missing consumption

```{r}
energy_growth <- energy %>%
  filter(year %in% c(2000, 2020)) %>%
  select(country, year, population, primary_energy_consumption) %>%
  filter(!is.na(population), !is.na(primary_energy_consumption)) 

```


Reshape to wide format: one row per country with columns for 2000 vs 2020

```{r}
energy_growth_wide <- energy_growth %>%
  pivot_wider(
    names_from = year,
    values_from = c(population, primary_energy_consumption),
    names_sep = "_"
  ) %>%
  # remove any countries for which 2000 or 2020 data is missing
  filter(!is.na(population_2000), !is.na(population_2020),
         !is.na(primary_energy_consumption_2000), !is.na(primary_energy_consumption_2020))

```


Computing absolute/percentage changes


```{r}
energy_growth_wide <- energy_growth_wide %>%
  mutate(
    pop_change = population_2020 - population_2000,
    energy_change = primary_energy_consumption_2020 - primary_energy_consumption_2000
    # pop_pct_change = (population_2020 - population_2000) / population_2000 * 100
    # energy_pct_change = (primary_energy_consumption_2020 - primary_energy_consumption_2000) /
    #                     primary_energy_consumption_2000 * 100
  )
```


Correlation & Visualization

```{r}
cor_test_result <- cor.test(energy_growth_wide$pop_change, energy_growth_wide$energy_change)
cor_test_result

# Scatter plot with a regression line
ggplot(energy_growth_wide, aes(x = pop_change, y = energy_change)) +
  geom_point(alpha = 0.5, color = "forestgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Population Growth vs. Energy Demand Growth (2000 - 2020)",
    x = "Change in Population",
    y = "Change in Primary Energy Consumption (TWh)"
  ) +
  theme_minimal()


# Log Transformed graph
ggplot(energy_growth_wide, aes(x = pop_change, y = energy_change)) +
  geom_point(alpha = 0.5, color = "forestgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_log10(labels = scales::comma) +  # Logarithmic scale for x-axis
  scale_y_log10(labels = scales::comma) +  # Logarithmic scale for y-axis
  labs(
    title = "Population Growth vs. Energy Demand Growth (2000 - 2020)",
    x = "Change in Population (Log Scale)",
    y = "Change in Primary Energy Consumption (Log Scale, TWh)"
  ) +
  theme_minimal()
```

